---
title: "GaugeRNA: Gauge-Invariant Audit of RNA-seq Differential Expression"
author: "GaugeRNA Research Group"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GaugeRNA: Gauge-Invariant Audit of RNA-seq Differential Expression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

GaugeRNA is a gauge-invariant audit framework for RNA-seq differential expression analysis. 
It quantifies the sensitivity of differential expression results to arbitrary normalization 
choices (gauge choices) through the Gauge Instability Score (GIS).

The key idea is that any normalization method implicitly chooses a "gauge" -- a reference 
frame for measuring expression. Different normalization choices can lead to different 
differential expression calls. GaugeRNA systematically audits the sensitivity of results 
to these choices.

### Key Concepts

1. **Gauge Instability Score (GIS)**: Quantifies how much differential expression results 
   change under random gauge perturbations. Lower GIS = more stable results.

2. **Gamma Projection**: Removes the gene-independent gauge component from log-fold changes 
   by centering around the median.

3. **Gauge Decomposition**: Factorizes the log-count matrix into gene gauge, sample gauge, 
   and curvature components.

4. **Audit Report**: A comprehensive report that combines all analyses and provides 
   actionable recommendations.

## 1. Basic Usage with Simulated Data

First, load the package and simulate some RNA-seq count data:

```{r simulate-data}
library(GaugeRNA)

# Set seed for reproducibility
set.seed(42)

# Simulate 500 genes x 10 samples
n_genes <- 500
n_samples <- 10

# Baseline expression from a negative binomial distribution
expr_mat <- matrix(rnbinom(n_genes * n_samples, mu = 100, size = 10), 
                   nrow = n_genes, ncol = n_samples)

# Add some true differential expression for 50 genes
de_genes <- sample(seq_len(n_genes), 50)
expr_mat[de_genes, 6:10] <- expr_mat[de_genes, 6:10] * 2

# Create condition vector (5 control, 5 treatment)
condition <- c(rep(0, 5), rep(1, 5))

# Add gene names
rownames(expr_mat) <- paste0("Gene", seq_len(n_genes))
colnames(expr_mat) <- paste0("Sample", seq_len(n_samples))

# View the data
dim(expr_mat)
head(expr_mat[, 1:6])
```

## 2. GIS Computation and Interpretation

The Gauge Instability Score measures the stability of differential expression results 
under random gauge transformations.

```{r compute-gis}
# Compute GIS for three methods
gis_results <- compute_gis(
  expr_mat = expr_mat,
  condition = condition,
  n_perturbations = 20,
  sigma = 1.0,
  methods = c("raw", "gamma", "deseq2")
)

# Display results
print(gis_results)

# Interpretation:
# - Lower GIS values indicate more gauge-stable methods
# - The gamma-projected method should generally have the lowest GIS
# - mean_stability close to 1 means the method is robust to gauge perturbations
```

The GIS values tell us:

- **raw**: The raw log-fold change method. May be sensitive to gauge choices.
- **gamma**: The gamma-projected method. Should be the most gauge-stable.
- **deseq2**: A DESeq2-like approximation. Provides a comparison against a 
  widely-used method.

```{r gis-interpretation}
# Find the best method
best_idx <- which.min(gis_results$GIS)
cat("Best method:", gis_results$method[best_idx], 
    "with GIS =", round(gis_results$GIS[best_idx], 4), "\n")

# Find the worst method
worst_idx <- which.max(gis_results$GIS)
cat("Worst method:", gis_results$method[worst_idx], 
    "with GIS =", round(gis_results$GIS[worst_idx], 4), "\n")
```

## 3. Gamma Projection

The gamma projection removes the arbitrary global shift from log-fold changes:

```{r gamma-projection}
# Compute raw log-fold changes
ctrl_idx <- which(condition == 0)
case_idx <- which(condition == 1)
ctrl_mean <- rowMeans(log2(expr_mat[, ctrl_idx] + 1))
case_mean <- rowMeans(log2(expr_mat[, case_idx] + 1))
raw_logfc <- case_mean - ctrl_mean

# Apply gamma projection
gamma_logfc <- gamma_projection(raw_logfc)

# Compare distributions
summary(raw_logfc)
summary(gamma_logfc)

# The median of gamma-projected logFC should be approximately 0
median(gamma_logfc)

# Plot comparison
plot(density(raw_logfc), main = "LogFC Distribution: Raw vs Gamma-Projected",
     xlab = "Log Fold Change", col = "blue", lwd = 2)
lines(density(gamma_logfc), col = "red", lwd = 2)
legend("topright", legend = c("Raw logFC", "Gamma-projected logFC"),
       col = c("blue", "red"), lwd = 2)
```

## 4. Gauge Decomposition

The gauge decomposition separates the expression matrix into interpretable components:

```{r gauge-decomposition}
# Perform gauge decomposition
decomp <- gauge_decomposition(expr_mat, pseudo_count = 1)

# Print summary
print(decomp)

# Examine the gene gauge (mu) - average expression per gene
summary(decomp$mu)

# Examine the sample gauge (a) - global shift per sample
summary(decomp$a)

# The curvature (epsilon) should have mean near 0
mean(as.vector(decomp$epsilon))

# Visualize the decomposition
par(mfrow = c(1, 3))
hist(decomp$mu, main = "Gene Gauge (mu)", xlab = "log2 expression", col = "steelblue")
hist(decomp$a, main = "Sample Gauge (a)", xlab = "log2 shift", col = "darkgreen")
hist(as.vector(decomp$epsilon), main = "Curvature (epsilon)", 
     xlab = "log2 residual", col = "darkred")
par(mfrow = c(1, 1))
```

```{r gauge-variance}
# Compare variance components
var_gene <- var(decomp$mu)
var_sample <- var(decomp$a)
var_curvature <- var(as.vector(decomp$epsilon))

cat("Variance decomposition:\n")
cat("  Gene gauge variance:", round(var_gene, 3), "\n")
cat("  Sample gauge variance:", round(var_sample, 3), "\n")
cat("  Curvature variance:", round(var_curvature, 3), "\n")
cat("  Total variance:", round(var_gene + var_sample + var_curvature, 3), "\n")
```

## 5. Audit Report Generation

The audit report is the main entry point for a comprehensive analysis:

```{r audit-report}
# Generate a complete audit report
report <- audit_report(
  expr_mat = expr_mat,
  condition = condition,
  n_perturbations = 15,
  sigma = 1.0,
  methods = c("raw", "gamma", "deseq2"),
  gis_threshold = "auto"
)

# Print the report
print(report)

# View the summary
summary(report)

# Examine the classification
head(report$classification)

# View recommendations
cat("\nRecommendations:\n")
for (i in seq_along(report$recommendations)) {
  cat(i, ". ", report$recommendations[i], "\n", sep = "")
}
```

```{r classification-summary}
# Summary of gene classification
class_table <- table(report$classification$classification)
barplot(class_table, 
        main = "Gene Classification: Gauge-Sensitive vs Gauge-Stable",
        col = c("darkred", "darkgreen"),
        ylab = "Number of Genes")
```

## 6. Using the Pre-Computed Data

The package includes pre-computed GIS results from the GaugeRNA paper:

```{r precomputed-data}
# Load pre-computed GIS results
data("gis_paper_results", package = "GaugeRNA")

# Examine the results
print(gis_paper_results)
```

## 7. Working with Real Data

For real RNA-seq datasets, the workflow is the same:

```{r real-data-example, eval=FALSE}
# Example workflow with a real dataset
# counts <- read.csv("gene_counts.csv", row.names = 1)
# condition <- c(rep(0, n_controls), rep(1, n_cases))
# 
# report <- audit_report(
#   expr_mat = as.matrix(counts),
#   condition = condition,
#   n_perturbations = 50  # Use more perturbations for real data
# )
# 
# print(report)
# write.csv(report$classification, "gauge_classification.csv")
```

## Conclusion

GaugeRNA provides a systematic framework for auditing the robustness of RNA-seq 
differential expression results. By quantifying gauge sensitivity, applying gamma 
projection, and decomposing expression matrices, researchers can:

1. Identify which differential expression methods are most robust to normalization choices
2. Flag genes whose differential expression calls are gauge-sensitive
3. Understand the variance structure of their data through gauge decomposition
4. Generate reproducible audit reports for publication

## Session Info

```{r session-info}
sessionInfo()
```